---
title: "Architecture and design philosophy"
format: html
vignette: >
  %\VignetteIndexEntry{Architecture and design philosophy}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: true
  echo: true
---

```{r}
#| label: setup
#| include: false
devtools::load_all("../")
```

## The problem

To customise a Quarto Word document today, you must:

1. Render a document
2. Open it in Microsoft Word
3. Manually click through "Modify Style" menus for *every single style*
4. Save this as `reference.docx`
5. Commit this binary blob to Git

**Why this fails:**

- **Opaque**: You cannot diff a binary file or see what changed between versions
- **Fragile**: Changing your brand colour requires manually editing 20+ styles
- **Unmaintainable**: There's no single source of truth—the style lives in a black box

## The solution: style-as-code

docstyle treats `reference.docx` as a **compiled artifact**, not an editable file.
Like a PDF or executable, you never edit it directly. Instead, you edit the source
code (CSS), and docstyle compiles a fresh `reference.docx` for you.

```
CSS (human-readable) → docstyle → reference.docx (artifact)
                                        ↓
                               Quarto/Pandoc → Final document
```

## CSS as lingua franca

CSS has evolved far beyond web styling. It's becoming the universal language for
visual specification across platforms:

| Domain | CSS role |
|--------|----------|
| **Web** | Native styling (original use case) |
| **Print/PDF** | `@page` rules, Paged.js, WeasyPrint |
| **Word/OOXML** | docstyle translates CSS → Word styles |
| **SVG** | Inline styles, external stylesheets, selectors for paths |
| **Desktop apps** | Electron, React Native, Flutter styling paradigms |
| **Design systems** | Design tokens often export to CSS variables |

The pattern: CSS provides a declarative, human-readable syntax. Tools translate
it to their native format (OpenXML for Word, PostScript for PDF, render trees
for native apps).

**Why this matters for docstyle:**

1. **Familiarity**: Most developers and designers already know CSS
2. **Tooling**: Syntax highlighting, linting, formatters, IDE support
3. **Diffability**: Text-based, easy to review changes in Git
4. **Composability**: Layered stylesheets, cascading rules, variables
5. **Future-proof**: CSS is a living standard with active W3C development

docstyle's investment in CSS support—`@page` rules, named page styles, custom
properties for line numbers—aligns with this broader trend. As CSS Paged Media
matures (via specs like CSS Page Floats and CSS Generated Content for Paged Media),
the same stylesheets may work across more output formats with less translation.

**The convergence hypothesis:** We may be moving toward a world where CSS is
the universal styling language, with format-specific renderers (browsers for HTML,
WeasyPrint/Prince for PDF, docstyle for Word) all consuming the same source.
This isn't fully realised today, but the trajectory is clear.

**Readability enables collaboration:** Markdown won for content because it's
readable. CSS is winning for styling for the same reason—people can read it,
edit it, and discuss it. In a world where AI assists with code, it can assist
with styling too. CSS provides a readable surface for that collaboration: the
human can review and refine what the AI generates, rather than trusting opaque
output like Word XML or LaTeX.

## Core principles

### 1. CSS as visual truth

The aesthetic definition of your brand—fonts, colours, sizes—exists in one place
only: the CSS file.

```css
/* popcorn-print.css - Print/Word styling */
h1 {
  font-family: "Libre Baskerville", "Georgia", serif;
  font-size: 14pt;
  font-weight: bold;
  color: #262c3a;
}

p {
  font-family: "Hanken Grotesk", "Arial", sans-serif;
  font-size: 11pt;
  color: #000000;
}
```

docstyle parses this CSS and translates it into the XML dialect Word requires
(`word/styles.xml`). If it's blue, it's defined in CSS.

### 2. Platform-specific styling (design tokens)

A brand needs consistent identity across platforms, but **appropriate expression**
for each medium. Web pages use large display fonts for impact; printed documents
use conservative typography for readability.

docstyle supports this through platform-specific CSS files:

```
popcorn.css         → Web (Basically A Mono headings, large scale)
popcorn-print.css   → Word/PDF (Libre Baskerville headings, traditional scale)
```
Both share:

- The colour palette (#262c3a Valhalla, #fdc882 Chardonnay)
- Body font (Hanken Grotesk)
- Visual hierarchy ratios

But differ in:

- Heading typeface (display vs serif)
- Font sizes (48px/32px web vs 14pt/13pt print)

This follows the **design token** pattern from modern design systems: abstract
values that translate appropriately per platform.

### 3. Layout truth from configuration

Document geometry (margins, headers, page size) comes from the project
configuration, typically `_quarto.yml`. This separates visual styling from
structural layout.

### 4. Clean slate base

We start from a minimal base document that contains the structural styles
Pandoc expects (Body Text, VerbatimChar, etc.). This skeleton is bundled with
docstyle and provides a stable foundation.

## The generation process

```
1. Copy Base Doc (Skeleton)
   └── Contains Pandoc's expected style structure

2. Inject Visual Styles (CSS)
   └── Translate CSS properties to OOXML <w:rPr> and <w:pPr>

3. Inject Page Layout (Optional)
   └── Margins, headers, footers from configuration

4. Save as reference.docx
   └── Ready for Quarto/Pandoc to use
```

## CSS to Word mapping

Not all CSS translates to Word, but docstyle supports the most important properties:

```{r}
#| label: mapping-table
#| echo: false
mapping <- data.frame(
  CSS = c("font-family", "font-size", "font-weight", "font-style", "color", "font (shorthand)"),
  Word = c("w:rFonts", "w:sz", "w:b", "w:i", "w:color", "(expands)"),
  Notes = c(
    "First font in stack",
    "Converted to half-points",
    "600+ = bold",
    "italic → italic",
    "Hex to OOXML",
    "Parses to components"
  )
)
knitr::kable(mapping, col.names = c("CSS Property", "Word XML", "Notes"))
```

### Unit conversion

Word uses its own unit system. docstyle handles the conversions:

```{r}
#| label: units
# Font sizes: CSS to Word half-points (1pt = 2 half-points)
css_to_half_points("16px")  # 16px
css_to_half_points("12pt")  # 12pt

# Margins: CSS to Word twips
css_to_twips("1in")         # 1 inch
```

## Semantic mapping for complex content

Standard markdown supports headings and paragraphs. For scientific documents,
you need abstracts, author affiliations, and custom callouts. docstyle uses
**semantic CSS mapping**:

1. **CSS**: Define `.abstract { font-style: italic; }`
2. **docstyle**: Creates "Abstract" style in Word
3. **QMD**: Use fenced divs: `::: {.abstract}`
4. **Pandoc**: Maps the div class to the Word style automatically

This keeps your `.qmd` semantic while centralising styling in CSS.

## Round-trip collaboration

docstyle enables collaboration with Word users without losing your markdown
source:

```
       You                    Collaborator
        │                          │
   Write .qmd                      │
        │                          │
   Render to .docx ──────────→  Edit in Word
        │                          │
        │              ←──────── Return .docx
        │                          │
   docx_to_qmd()                   │
        │                          │
   Review diff in Git              │
        │                          │
   Merge changes                   │
```

Citations from Zotero are extracted and converted to Quarto format, so the
round-trip preserves bibliographic data.

## Implementation notes

### XML injection

docstyle uses the `officer` package to read/write Word documents and `xml2`
for direct XML manipulation. Style definitions live in `word/styles.xml`
within the docx container (which is just a ZIP file).

```{r}
#| label: xml-demo
# Generate a reference document dynamically from config
config_path <- system.file("extdata/minimal-example/_quarto.yml", package = "docstyle")
if (config_path == "") config_path <- "../inst/extdata/minimal-example/_quarto.yml"

ref_path <- tempfile(fileext = ".docx")
generate_reference_doc(config_path, output_path = ref_path)

doc <- officer::read_docx(ref_path)

# Access the styles XML
styles_xml <- get_styles_xml(doc)

# Find a specific style
ns <- c(w = "http://schemas.openxmlformats.org/wordprocessingml/2006/main")
h1 <- xml2::xml_find_first(styles_xml, "//w:style[@w:styleId='Heading1']", ns)

# Show the run properties (font, size, colour)
rPr <- xml2::xml_find_first(h1, ".//w:rPr", ns)
as.character(rPr)
```

### Why not use Pandoc directly?

Pandoc already handles markdown → Word conversion. docstyle doesn't replace
Pandoc—it manufactures the `reference.docx` that Pandoc needs. This separation
keeps each tool focused on what it does best:

- **Pandoc**: Content conversion (markdown → Word paragraphs)
- **docstyle**: Style management (CSS → Word styles)

## Comparison with alternatives

| Approach | Pros | Cons |
|----------|------|------|
| Manual Word editing | Familiar | No version control, error-prone |
| YAML-only config | Simple syntax | Limited expressiveness |
| CSS (docstyle) | Web-standard, diffable, powerful | Requires CSS knowledge |

docstyle is best for teams who already use CSS for web styling and want to
maintain consistency across HTML, PDF, and Word outputs.

## Further reading

**Tutorials:**

- [Getting started](tutorial-getting-started.html): Quick introduction
- [Round-trip workflow](how-to-roundtrip.html): Word ↔ Quarto collaboration

**How-to guides:**

- [CSS to Word](css-to-word.html): Create templates from CSS
- [Word to Quarto](word-to-quarto.html): Convert documents back
- [Comments and revisions](comments-revisions.html): Track changes workflow
- [Zotero citations](zotero-citations.html): Citation handling

**Explanation:**

- [Understanding Word's format](explanation-word-xml.html): How DOCX files work internally
- [Tables: An evolving ecosystem](explanation-tables.html): Table packages and formats

**Reference:**

- [YAML schema](reference-yaml-schema.html): All configuration options
- [File conventions](reference-file-conventions.html): Directory structure and sidecar files
