---
title: "Comments and revisions round-trip"
format: html
vignette: >
  %\VignetteIndexEntry{Comments and revisions round-trip}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: true
  echo: true
---

```{r}
#| label: setup
#| include: false
devtools::load_all("../")
```

## Overview

The comments and revisions workflow enables round-tripping Word comments and track changes (insertions/deletions) through Quarto editing. This allows collaborative review workflows where:

1. Colleagues add comments and track changes in Word
2. You import to QMD for editing in a plain-text environment
3. Comments and revisions are preserved as lightweight spans
4. You render back to Word with full comment/revision metadata intact

**Key design:** docstyle uses a "hybrid span + sidecar" model—comments and revisions appear as simple Pandoc spans in your QMD, while rich metadata (author, date, initials) lives in JSON sidecar files.

## The workflow

```
┌─────────────────┐     extract_comments()    ┌─────────────────┐
│   Word (.docx)  │     extract_revisions()   │   Quarto (.qmd) │
│   with comments │ ──────────────────────▶   │   + JSON files  │
│   and changes   │     docx_to_qmd()         │                 │
└─────────────────┘                           └─────────────────┘
                                                      │
                                                      │ edit in Quarto
                                                      ▼
┌─────────────────┐     quarto render          ┌─────────────────┐
│   Word (.docx)  │     inject_comments()     │   Quarto (.qmd) │
│   comments and  │ ◀──────────────────────   │   edited version│
│   changes intact│                           │                 │
└─────────────────┘                           └─────────────────┘
```

## QMD syntax

### Comments

docstyle supports two formats for comment markers:

**Span-based (simple cases):**

```markdown
This paragraph contains [a comment on this phrase]{.comment id="3"} to demonstrate
the comment round-trip workflow.
```

**HTML markers (complex structures):**

For comments that span tracked changes, multiple formatting regions, or other complex structures, docstyle uses raw HTML markers:

```markdown
This `<!-- comment:start id="3" -->`{=html}phrase spans [inserted text]{.ins id="rev_1"} and more`<!-- comment:end id="3" -->`{=html} content.
```

The HTML marker format is used automatically during import when comments cross structural boundaries that Pandoc spans cannot represent. Both formats render identically to Word.

**Fallback behaviour:** If a start marker has no matching end marker (e.g., due to editing), docstyle renders it as a "point comment"—a comment marker at the start position with no range. Orphan end markers are silently ignored.

The `id` links to metadata in `comments.json`:

```json
{
  "3": {
    "id": "3",
    "author": "Reviewer 1",
    "date": "2025-01-15T10:30:00Z",
    "content": "Is this phrase clear enough for the target audience?",
    "parent_id": null,
    "initials": "R1"
  }
}
```

### Track changes: insertions

Insertions use spans with the `.ins` class:

```markdown
This paragraph demonstrates insertions: [this text was added]{.ins id="rev_7"} by a reviewer.
```

### Track changes: deletions

Deletions combine strikethrough with the `.del` class:
```markdown
This paragraph demonstrates deletions: [~~this text was removed~~]{.del id="rev_8"} by an editor.
```

Both insertions and deletions link to metadata in `revisions.json`:

```json
{
  "rev_7": {
    "id": "rev_7",
    "type": "insertion",
    "author": "Jane Doe",
    "date": "2025-01-20T14:00:00Z"
  },
  "rev_8": {
    "id": "rev_8",
    "type": "deletion",
    "author": "Editor",
    "date": "2025-01-21T09:30:00Z"
  }
}
```

## Step 1: Extract from Word

### Extract comments

```{r}
#| label: extract-comments-example
#| eval: false
# Extract comments from a Word document
comments <- extract_comments("reviewed-draft.docx")

# Write to JSON sidecar
write_comments_json(comments, "comments.json")

# What you get:
names(comments)
# [1] "1" "2" "3"  # Comment IDs

comments[["1"]]
# $id: "1"
# $author: "Reviewer 1"
# $date: "2025-01-15T10:30:00Z"
# $content: "Is this phrase clear?"
# $initials: "R1"
```

### Extract revisions (track changes)

```{r}
#| label: extract-revisions-example
#| eval: false
# Extract track changes from a Word document
revisions <- extract_revisions("reviewed-draft.docx")

# Write to JSON sidecar
write_revisions_json(revisions, "revisions.json")

# Summary of changes
count_revisions(revisions)
# insertions  deletions
#          5          3

summarise_revisions_by_author(revisions)
#        author insertions deletions
# 1    Jane Doe          3         1
# 2 Reviewer 2           2         2
```
### Convert document to QMD

The `docx_to_qmd()` function automatically detects comment ranges and track changes:

```{r}
#| label: convert-example
#| eval: false
# Convert Word to QMD with comment/revision spans
docx_to_qmd(
  "reviewed-draft.docx",
  output_path = "reviewed-draft.qmd"
)
```

The resulting QMD contains spans like:

```markdown
The committee [recommends]{.comment id="12"} that the [updated]{.ins id="rev_13"}
guidelines be [~~immediately~~]{.del id="rev_14"} adopted
[within the next quarter]{.ins id="rev_15"}.
```

## Step 2: Edit in Quarto

### Editing workflow

In your QMD, you interact with comments and revisions as plain text:

| User intent | Action in QMD | Result |
|-------------|---------------|--------|
| **Resolve comment** | Remove `[...]{.comment}` wrapper, keep text | Comment disappears |
| **Accept insertion** | Remove `[...]{.ins}` wrapper, keep text | Text becomes normal |
| **Reject insertion** | Delete the entire span | Text is removed |
| **Accept deletion** | Delete the entire span including text | Text is removed |
| **Reject deletion** | Remove `[~~...~~]{.del}` wrapper, keep text | Text is restored |
| **Defer decision** | Leave syntax as is | Re-appears in Word |

### Visual preview

When previewing your QMD as HTML, the CSS styles highlight comments and revisions:

- **Comments**: Yellow background with gold underline
- **Insertions**: Green background with underline
- **Deletions**: Red background with strikethrough

These styles are defined in your CSS (e.g., `popcorn-print.css`):

```css
.comment {
  background-color: #fff3cd;
  border-bottom: 2px solid #ffc107;
}

.ins {
  background-color: #e6ffec;
  text-decoration: underline;
}

.del {
  background-color: #ffebe9;
  text-decoration: line-through;
  color: #6a737d;
}
```

## Step 3: Render back to Word

### Basic rendering

```bash
# Render QMD back to Word
quarto render reviewed-draft.qmd
```

The Lua filters automatically convert spans to OpenXML:

- `.comment` spans → `w:commentRangeStart`, `w:commentRangeEnd`, `w:commentReference`
- `.ins` spans → `w:ins` with author/date
- `.del` spans → `w:del` with `w:delText`

### Injecting comments

After rendering, inject the `comments.xml` file:

```{r}
#| label: inject-comments
#| eval: false
# Inject comments into the rendered Word document
inject_comments(
  docx_path = "output/reviewed-draft.docx",
  comments_json = "comments.json"
)
```

This:

1. Scans the document for used comment IDs
2. Builds `word/comments.xml` with only the referenced comments

3. Updates `[Content_Types].xml` and `document.xml.rels`
4. Repacks the DOCX

### Passing revision metadata

For revisions to include author/date metadata, pass the JSON path via Pandoc metadata:

```{r}
#| label: render-with-revisions
#| eval: false
# In your _quarto.yml or render call:
# -M revisions-file:revisions.json

# Or programmatically:
system(paste(

  "quarto render reviewed-draft.qmd",
  "-M revisions-file:revisions.json",
  "--to docx"
))
```

## Complete example

Here's a full workflow using the test files included with docstyle:

```{r}
#| label: full-example
#| eval: false
# 1. Start with a Word document containing comments and track changes
input_docx <- "colleague-review.docx"

# 2. Extract comments and revisions
comments <- extract_comments(input_docx)
revisions <- extract_revisions(input_docx)

write_comments_json(comments, "comments.json")
write_revisions_json(revisions, "revisions.json")

# 3. Convert to QMD (spans are automatically inserted)
docx_to_qmd(input_docx, output_path = "colleague-review.qmd")

# 4. Edit in your favourite text editor...
# - Accept/reject changes by editing spans
# - Resolve comments by removing span markers
# - Add new content

# 5. Render back to Word
# quarto render colleague-review.qmd

# 6. Inject comments into the output
inject_comments(
  docx_path = "output/colleague-review.docx",
  comments_json = "comments.json"
)

# Result: Word document with preserved comments and track changes
```

## File structure

A typical project with comments/revisions:

```
project/
├── manuscript.qmd           # Contains [text]{.comment id="x"}, etc.
├── comments.json            # Comment metadata (author, date, content)
├── revisions.json           # Track changes metadata (author, date, type)
├── references.json          # CSL-JSON citations (if using Zotero)
├── field-codes.json         # Zotero field codes (if using Zotero)
├── _quarto.yml              # Project configuration
└── output/
    └── manuscript.docx      # Generated Word doc with comments
```

## Known limitations

### Comments spanning complex structures

Comments that span tracked changes, multiple paragraphs, or nested formatting are handled using HTML markers (see above). This approach is robust but makes the QMD slightly less readable:

```markdown
`<!-- comment:start id="3" -->`{=html}text with [insertion]{.ins id="rev_1"}`<!-- comment:end id="3" -->`{=html}
```

### Nested/overlapping comments

Word allows overlapping comment ranges (e.g., Comment A on words 1-3, Comment B on words 2-4). docstyle supports this using multiple comment markers, but deeply nested overlaps may require manual adjustment after import.

### Orphan markers

If you accidentally delete one end of a comment marker pair:

- **Missing end marker**: Renders as a point comment (marker at start position, no range)
- **Missing start marker**: Orphan end is silently removed

This fallback prevents rendering errors but may not match original intent.

### Formatting inside deletions

Rich formatting (bold, italic) inside deleted text is converted to plain text. The deletion itself is preserved, but internal formatting is lost.

### Move tracking

Word's "move" tracking (`w:moveFrom`/`w:moveTo`) is downgraded to simple deletion/insertion pairs.

## Troubleshooting

### Comments not appearing in Word

1. Ensure you called `inject_comments()` after rendering
2. Check that `comments.json` exists and contains valid data
3. Verify the comment IDs in your QMD match the JSON keys

### "Unknown" author on revisions

If revisions show "Unknown" as the author:

1. Ensure `revisions.json` exists with metadata
2. Pass the JSON path via `-M revisions-file:revisions.json`
3. Check that revision IDs match (the QMD uses `rev_X` format)

### Orphaned brackets in QMD

If you see stray `[` or `]` characters after import:

1. These likely come from complex comment structures
2. Check for mismatched HTML markers: `` `<!-- comment:start -->`{=html} `` without matching end
3. docstyle will render orphan starts as point comments, so the document will still build

### Render warnings about orphan comments

If you see warnings like:

```
[comment-inject] Warning: orphan start marker id=3 (will appear as point comment)
```

This means a comment start marker has no matching end. The comment will appear in Word at the start position without highlighting a text range. To fix:

1. Find the start marker in your QMD
2. Either add the missing end marker or remove the start marker entirely

## Further reading

- [Understanding Word's format](explanation-word-xml.html): Why round-tripping is challenging
- [File conventions](reference-file-conventions.html): Sidecar file formats
- [Round-trip workflow](how-to-roundtrip.html): General Word ↔ Quarto workflow
- [Zotero citations](zotero-citations.html): Combining comments with citation round-trip
- [Architecture](architecture.html): Technical details of the Lua filters and XML injection
