---
title: "Creating Word templates from CSS"
format: html
vignette: >
  %\VignetteIndexEntry{Creating Word templates from CSS}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: true
  echo: true
---

```{r}
#| label: setup
#| include: false
devtools::load_all("../")
```

## Overview

This guide shows how to create a Word `reference.docx` template from your own CSS
file. This is useful when you have:

- Brand guidelines defined in CSS
- A web style guide you want to match in Word output
- Custom styles beyond what Pandoc provides by default

## Writing CSS for Word

Not all CSS properties translate to Word. docstyle supports:

| CSS Property | Word Equivalent | Notes |
|--------------|-----------------|-------|
| `font-family` | `w:rFonts` | First font in stack used |
| `font-size` | `w:sz` | Converted to half-points |
| `font-weight` | `w:b` | 600+ treated as bold |
| `font-style` | `w:i` | `italic` → italic |
| `color` | `w:color` | Hex converted to OOXML |
| `font` | (shorthand) | Expands to above properties |

### Example CSS

```css
/* Headings - use your brand font */
h1 {
  font-family: "Georgia", serif;
  font-size: 24pt;
  font-weight: bold;
  color: #1a1a2e;
}

h2 {
  font-family: "Georgia", serif;
  font-size: 18pt;
  font-weight: bold;
  color: #1a1a2e;
}

/* Body text */
p {
  font-family: "Arial", sans-serif;
  font-size: 11pt;
  color: #333333;
}

/* Custom class for callouts */
.callout {
  font-family: "Arial", sans-serif;
  font-size: 10pt;
  font-style: italic;
  color: #666666;
}
```

## Creating a minimal CSS file

Let's create a simple CSS file and generate a template from it:

```{r}
#| label: create-css
# Create a temp directory for our example
example_dir <- file.path(tempdir(), "css-example")
dir.create(example_dir, showWarnings = FALSE)

# Create a minimal CSS file
css_content <- '
/* Minimal brand styles */
h1 {
  font-family: "Times New Roman", serif;
  font-size: 18pt;
  font-weight: bold;
  color: #000080;
}

h2 {
  font-family: "Times New Roman", serif;
  font-size: 14pt;
  font-weight: bold;
  color: #000080;
}

p {
  font-family: "Calibri", sans-serif;
  font-size: 11pt;
  color: #333333;
}
'

css_file <- file.path(example_dir, "my-brand.css")
writeLines(css_content, css_file)
cat("Created CSS file:", css_file, "\n")
```

## Creating a configuration file

docstyle uses a YAML configuration to specify the CSS file. Page settings are configured via `@page` rules in the CSS file itself:

```{r}
#| label: update-css-with-page
# Add @page rules to our CSS
css_with_page <- '
/* Page layout */
@page {
  size: letter;
  margin: 1in;
}

/* Minimal brand styles */
h1 {
  font-family: "Times New Roman", serif;
  font-size: 18pt;
  font-weight: bold;
  color: #000080;
}

h2 {
  font-family: "Times New Roman", serif;
  font-size: 14pt;
  font-weight: bold;
  color: #000080;
}

p {
  font-family: "Calibri", sans-serif;
  font-size: 11pt;
  color: #333333;
}
'

writeLines(css_with_page, css_file)
```

```{r}
#| label: create-config
config_content <- '
docstyle:
  css: my-brand.css
'

config_file <- file.path(example_dir, "_quarto.yml")
writeLines(config_content, config_file)
cat("Created config:", config_file, "\n")
```

## Generating the template

```{r}
#| label: generate-template
output_path <- file.path(example_dir, "reference.docx")
ref_doc <- generate_reference_doc(config_file, output_path = output_path)

cat("Generated:", basename(ref_doc), "\n")
cat("Size:", file.size(ref_doc), "bytes\n")
```

## Verifying the output

Let's confirm the styles were correctly injected:

```{r}
#| label: verify-output
doc <- officer::read_docx(ref_doc)
styles_xml <- get_styles_xml(doc)
ns <- c(w = "http://schemas.openxmlformats.org/wordprocessingml/2006/main")

# Check each heading level
for (level in 1:2) {
  style_id <- paste0("Heading", level)
  node <- xml2::xml_find_first(
    styles_xml,
    sprintf("//w:style[@w:styleId='%s']", style_id),
    ns
  )

  if (!inherits(node, "xml_missing")) {
    color <- xml2::xml_attr(xml2::xml_find_first(node, ".//w:color", ns), "val")
    size <- xml2::xml_attr(xml2::xml_find_first(node, ".//w:sz", ns), "val")
    cat(sprintf("Heading %d: colour=%s, size=%s half-points\n", level, color, size))
  }
}
```

The colour `000080` is navy blue, and the sizes (36 and 28 half-points) correspond
to 18pt and 14pt respectively.

## Using shorthand font property

CSS shorthand `font:` is also supported:

```{r}
#| label: shorthand-demo
# Parse a font shorthand
font_str <- '600 16px "Helvetica", "Arial", sans-serif'
parts <- parse_font_shorthand(font_str)
print(parts)
```

This extracts:

- `weight`: 600 (semi-bold)
- `size`: 16px
- `family`: "Helvetica", "Arial", sans-serif

## Unit conversion

docstyle handles CSS units and converts them to Word's internal units:

```{r}
#| label: unit-conversion
# CSS pixels to Word half-points
# Word uses half-points for font sizes (1pt = 2 half-points)
# CSS standard: 1px = 0.75pt
css_to_half_points("16px")  # 16px → half-points
css_to_half_points("12pt")  # 12pt → half-points

# CSS inches to Word twips (for margins)
css_to_twips("1in")         # 1 inch → twips
```

## CSS selector mapping

Selectors map to Word style IDs:

| CSS Selector | Word Style ID | Word Style Name |
|--------------|---------------|-----------------|
| `h1` | Heading1 | Heading 1 |
| `h2` | Heading2 | Heading 2 |
| `p` | Normal | Normal |
| `.custom` | custom | Custom |

Custom classes (`.name`) create new Word styles with that name.

## Next steps

- **[CSS property reference](reference-css-properties.html)**: Complete list of supported CSS properties
- **[Getting started](tutorial-getting-started.html)**: Full workflow tutorial
- **[Zotero citations](zotero-citations.html)**: Handle citations in the round-trip
- **[Architecture](architecture.html)**: Understand the design decisions
