---
title: "Getting started with docstyle"
format: html
vignette: >
  %\VignetteIndexEntry{Getting started with docstyle}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: true
  echo: true
---

```{r}
#| label: setup
#| include: false
devtools::load_all("../")
```

## Overview

**docstyle** is a Quarto extension that generates styled Word documents from CSS + YAML configuration. Define your styles once in CSS, configure page layout in YAML, and let docstyle generate the `reference.docx` automatically.

This tutorial walks through the core workflow: **QMD → Word**.

## The problem

Microsoft Word's `reference.docx` templates are:

-   **Fragile**: Small edits can break styles in unexpected ways
-   **Opaque**: Hard to diff, review, or version control
-   **Manual**: Every style change requires editing the Word file

## The solution

docstyle uses CSS + YAML as the single source of truth:

```
styles.css + _quarto.yml → reference.docx → Quarto → Final .docx
```

The extension generates `reference.docx` automatically before each render.

## Installation

docstyle requires **both** the Quarto extension and the R package.

### Step 1: Install the R package

```r
remotes::install_github("DougManuel/docstyle")
```

The R package provides the CSS parsing and reference document generation that runs before each render.

### Step 2: Install the Quarto extension

**Option A: From R (recommended)**

```r
docstyle::use_docstyle()
```

This installs the extension from the package, ensuring version consistency.

**Option B: From GitHub**

```bash
quarto add DougManuel/docstyle
```

Both methods create `_extensions/docstyle/` in your project.

### Verify installation

After installation, you should have:

- `_extensions/docstyle/` directory with Lua filters and R scripts
- The `docstyle` R package available (`library(docstyle)` should work)

### Subdirectory projects

If your QMD files are in subdirectories (e.g., `reports/annual-report/`), Quarto looks for `_extensions/` relative to the QMD file. You need to make the extension accessible:

**Option A: Symlink (recommended)**

```bash
cd reports/annual-report
mkdir -p _extensions
ln -s ../../_extensions/docstyle _extensions/docstyle
```

**Option B: Install in subdirectory**

```bash
cd reports/annual-report
quarto add DougManuel/docstyle
```

The bootstrap scripts in `_docstyle/` handle extension auto-restore, so subdirectory projects work without path adjustments.

See the [Troubleshooting guide](troubleshooting.html) for detailed subdirectory setup instructions.

## Quick start

### Step 1: Create a CSS file

Create `styles.css` with page layout and typography:

```css
/* Page layout */
@page {
  size: letter;
  margin: 1in;
}

/* Named page style for landscape sections */
@page landscape {
  size: letter landscape;
  margin: 0.5in;
}

/* Headings */
h1 {
  font: 600 36px "Arial";
  color: #262C3A;
}

h2 {
  font: 600 24px "Arial";
  color: #262C3A;
}

/* Body text */
p {
  font: 400 11px "Arial";
  line-height: 1.5;
  color: #333333;
}
```

### Step 2: Configure `_quarto.yml`

```yaml
project:
  type: default

format:
  docstyle-docx:
    docstyle:
      css: styles.css
      footer:
        enabled: true
        right: "Page {page} of {pages}"
```

### Step 3: Write your document

Create `document.qmd`:

```markdown
---
title: "My Document"
---

# Introduction

This is body text styled by CSS.

## Methods

Another section with consistent styling.
```

### Step 4: Render

```bash
quarto render document.qmd
```

The extension:

1. Reads your CSS and YAML configuration
2. Generates a `reference.docx` with those styles
3. Renders your QMD using that reference document

## Live example

Let's see docstyle in action using the bundled POPCORN study styles.

### The CSS

```{r}
#| label: read-css
css_path <- system.file("extdata/popcorn/popcorn.css", package = "docstyle")
styles <- read_css(css_path)
```

**h1 style:**

```{r}
#| label: show-h1
styles[["h1"]]
```

**p (body) style:**

```{r}
#| label: show-p
styles[["p"]]
```

### Generate a reference document

```{r}
#| label: generate-reference
#| results: hide
# Create a config
config <- list(
  docstyle = list(
    css = css_path,
    page = list(
      size = "letter",
      margins = list(top = "1in", bottom = "1in", left = "1in", right = "1in")
    ),
    footer = list(
      enabled = TRUE,
      right = "Page {page} of {pages}"
    )
  )
)

output_path <- file.path(tempdir(), "reference.docx")
ref_doc <- generate_reference_doc(config, output_path = output_path)

cat("Generated:", ref_doc, "\n")
cat("File size:", file.size(ref_doc), "bytes\n")
```

### Verify the styles

```{r}
#| label: verify-styles
doc <- officer::read_docx(ref_doc)
styles_xml <- get_styles_xml(doc)
ns <- c(w = "http://schemas.openxmlformats.org/wordprocessingml/2006/main")

# Check Heading 1
h1_node <- xml2::xml_find_first(styles_xml, "//w:style[@w:styleId='Heading1']", ns)
h1_color <- xml2::xml_find_first(h1_node, ".//w:color", ns)
h1_sz <- xml2::xml_find_first(h1_node, ".//w:sz", ns)

cat("Heading 1 colour:", xml2::xml_attr(h1_color, "val"), "\n")
cat("Heading 1 size (half-points):", xml2::xml_attr(h1_sz, "val"), "\n")
```

The colour `262C3A` is POPCORN's dark blue, and 72 half-points = 36pt (matching the CSS).

## Configuration reference

Page layout is configured in CSS via `@page` rules:

```css
@page {
  size: letter;            /* letter | a4 | legal */
  margin: 1in;             /* shorthand or individual margins */

  /* Optional: line numbers */
  --docstyle-line-numbers: every 1;
  --docstyle-line-numbers-restart: page;
}
```

The `docstyle:` section in YAML handles headers, footers, and metadata:

```yaml
docstyle:
  css: styles.css           # Path to CSS file (includes @page rules)

  # Footer
  footer:
    enabled: true
    content: "Page {page}"  # Single-position format
    align: center
    # Or multi-position format:
    left: "Document Title"
    center: ""
    right: "Page {page} of {pages}"
    style: footer           # CSS class for styling
    first-page: false       # Different first page
```

## Configuration tips

### Where to put configuration

docstyle configuration is split between two locations:

| Configuration | Location | Used by |
|--------------|----------|---------|
| `css:`, `footer:`, `header:` | `_quarto.yml` | Pre-render R script |
| `author-plate:`, `authors:`, `affiliations:` | QMD front matter | Lua filters at render time |

Lua filter configuration (like `author-plate:`) must be in the QMD front matter because Quarto strips nested `docstyle.*` keys from `_quarto.yml` before passing to filters.

See the [Author plates guide](author-plate.html) for detailed configuration.

## Next steps

- **[Troubleshooting](troubleshooting.html)**: Common issues and solutions
- **[CSS to Word](css-to-word.html)**: Detailed guide on CSS property mapping
- **[Author plates](author-plate.html)**: Generate formatted author blocks for manuscripts
- **[Round-trip workflow](how-to-roundtrip.html)**: Collaborate with Word users (advanced)
- **[Zotero citations](zotero-citations.html)**: Handle citations in Word ↔ QMD workflows
