---
title: "Converting Word documents to Quarto"
format: html
vignette: >
  %\VignetteIndexEntry{Converting Word documents to Quarto}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: true
  echo: true
---

```{r}
#| label: setup
#| include: false
devtools::load_all("../")
```

## Overview

This guide shows how to convert existing Word documents to Quarto markdown. This is
useful when:

- Collaborators have edited a Word document you need to bring back to Quarto
- You're starting from an existing Word-based workflow
- You want to extract and preserve document structure

## Creating a test document

First, let's create a simple Word document to convert. In practice, you would use
an existing document from collaborators.

```{r}
#| label: create-test-doc
# Create a simple Word document using officer
library(officer)

# Generate a reference document from the bundled minimal example.
config_path <- system.file(
  "extdata/minimal-example/_quarto.yml",
  package = "docstyle"
)

base_docx <- tempfile(fileext = ".docx")
generate_reference_doc(config_path, output_path = base_docx)

doc <- read_docx(base_docx)

# Clear existing content and add our own
doc <- body_add_par(doc, "Sample Research Report", style = "Title")

# Add content
doc <- body_add_par(doc, "Introduction", style = "Heading 1")
doc <- body_add_par(doc,
  "This document demonstrates the Word to Quarto conversion process. The docstyle package preserves heading structure, paragraph text, and document metadata.",
  style = "Normal"
)

doc <- body_add_par(doc, "Methods", style = "Heading 1")
doc <- body_add_par(doc, "Data collection", style = "Heading 2")
doc <- body_add_par(doc,
  "We collected data using standardised protocols across all study sites.",
  style = "Normal"
)

doc <- body_add_par(doc, "Analysis approach", style = "Heading 2")
doc <- body_add_par(doc,
  "Statistical analysis was performed using R version 4.3.0.",
  style = "Normal"
)

doc <- body_add_par(doc, "Results", style = "Heading 1")
doc <- body_add_par(doc,
  "The analysis revealed significant findings across all measures.",
  style = "Normal"
)

# Save to temp file
test_docx <- file.path(tempdir(), "test-report.docx")
print(doc, target = test_docx)

cat("Created test document:", test_docx, "\n")
cat("Size:", file.size(test_docx), "bytes\n")
```

## Converting to Quarto

The `docx_to_qmd()` function handles the conversion:

```{r}
#| label: convert-doc
output_qmd <- file.path(tempdir(), "test-report.qmd")

# Convert the document
result <- docx_to_qmd(
  docx_path = test_docx,
  output_path = output_qmd
)

cat("Created:", result, "\n")
```

## Examining the output

Let's see what the conversion produced:

```{r}
#| label: show-output
qmd_content <- readLines(output_qmd)
cat(qmd_content, sep = "\n")
```

The conversion:

- Extracted the title to YAML front matter
- Converted Word heading styles to markdown headings (`#`, `##`)
- Preserved paragraph text
- Set up the document format as `docx` for round-tripping

## Using a style template

You can apply a docstyle template during conversion:

```{r}
#| label: convert-with-style
output_styled <- file.path(tempdir(), "test-report-styled.qmd")

# Convert with POPCORN style
result_styled <- docx_to_qmd(
  docx_path = test_docx,
  output_path = output_styled,
  style = "popcorn"
)

# Show the YAML header
styled_content <- readLines(output_styled)
yaml_end <- which(styled_content == "---")[2]
cat(styled_content[1:yaml_end], sep = "\n")
```

The `reference-doc` field is automatically added, so rendering back to Word will
use consistent styling.

## Style mapping

Word styles are mapped to markdown elements:

| Word Style | Markdown |
|------------|----------|
| Title | YAML `title:` |
| Subtitle | YAML `subtitle:` |
| Heading 1 | `# Heading` |
| Heading 2 | `## Heading` |
| Heading 3 | `### Heading` |
| ListBullet | `- item` |
| ListNumber | `1. item` |
| Normal | Paragraph text |

## Handling lists

List items are converted to markdown lists:

```{r}
#| label: list-example
# Create doc with lists (using POPCORN template for styles)
doc_lists <- read_docx(base_docx)
doc_lists <- body_add_par(doc_lists, "Project Tasks", style = "Title")
doc_lists <- body_add_par(doc_lists, "Key deliverables:", style = "Normal")
doc_lists <- body_add_par(doc_lists, "Complete data analysis", style = "Block Text")
doc_lists <- body_add_par(doc_lists, "Write methods section", style = "Block Text")
doc_lists <- body_add_par(doc_lists, "Prepare figures", style = "Block Text")

list_docx <- file.path(tempdir(), "list-example.docx")
print(doc_lists, target = list_docx)

# Convert
list_qmd <- docx_to_qmd(list_docx)
cat(readLines(list_qmd), sep = "\n")
```

## Round-trip workflow

The typical collaboration workflow is:

1. **Start in Quarto**: Write your document in `.qmd`
2. **Render to Word**: `quarto render doc.qmd --to docx`
3. **Collaborator edits**: Send `.docx` for review/editing
4. **Convert back**: `docx_to_qmd("edited.docx")`
5. **Diff and merge**: Review changes in your `.qmd`

This preserves the benefits of version-controlled markdown while allowing
Word-based collaboration.

## Preserving YAML headers on re-import

When re-importing an edited Word document, you often want to keep your curated
YAML front matter (author info, affiliations, version history) while getting
the updated body content. The `preserve_header` parameter handles this:

```{r}
#| label: preserve-header-example
#| eval: false
# Re-import edited document, keeping your existing YAML header
docx_to_qmd(
  docx_path = "edited-by-colleague.docx",
  output_path = "manuscript.qmd",
  preserve_header = TRUE  # This is the default
)
```

When `preserve_header = TRUE` and the output file already exists:

1. The existing YAML header is extracted and preserved

2. The old file is renamed to `*_old.qmd` as a backup
3. The new file gets: preserved header + fresh body content from Word

This is particularly useful when you've added rich metadata like:

```yaml
---
title: "My Research Paper"
author:
  - name: "Jane Smith"
    orcid: "0000-0001-2345-6789"
    affiliations:
      - ref: uottawa
version: "1.2"
---
```

Set `preserve_header = FALSE` to regenerate the YAML header from the Word document's
Title/Subtitle styles.

## Next steps

- **[Zotero citations](zotero-citations.html)**: Handle citation extraction during conversion
- **[CSS to Word](css-to-word.html)**: Create custom templates from CSS
- **[Architecture](architecture.html)**: Understand the design philosophy
