---
title: "Working with Zotero citations"
format: html
vignette: >
  %\VignetteIndexEntry{Working with Zotero citations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
  echo: true
---

```{r}
#| label: setup
#| include: false
#| eval: true
devtools::load_all("../")
```

## Overview

Collaborating with Word users who use Zotero can be challenging. When you convert a Word document to Markdown, citations usually become static text. When you convert back to Word, they remain static—breaking your collaborator's workflow.

**docstyle** solves this by preserving live, editable Zotero citations through the round-trip.

## Quick start

The simplest workflow:

```{r}
#| label: quick-start
library(docstyle)

# 1. Import Word document (extracts content + citations)
result <- import_docx("colleague-draft.docx")

# 2. Edit the QMD file...

# 3. Render back to Word with live Zotero citations
# quarto render colleague-draft.qmd
```

That's it. Your colleague receives a Word document where Zotero citations work normally.

## How it works

### JSON-based citation management

Unlike typical Quarto workflows that use `.bib` files, docstyle keeps reference metadata in JSON format—similar to how Zotero manages citations in Word:

| File | Purpose |
|------|---------|
| `references.json` | CSL-JSON citation data for rendering |
| `field-codes.json` | Zotero field code metadata for injection |

This approach:

- Avoids merge conflicts when multiple authors edit citations
- Keeps metadata "behind the scenes" during editing
- Enables seamless round-trip with Word

### The three-step process

1. **Extract**: `import_docx()` reads Zotero field codes from Word and saves them as JSON
2. **Edit**: Work in Quarto using standard `[@key]` citation syntax
3. **Render**: `quarto render` with the docstyle extension injects field codes back into the output Word document

## Detailed workflow

### Step 1: Extract from Word

When you receive a Word document with Zotero citations:

```{r}
#| label: extract
# Extract citations and content in one step
result <- import_docx(
  docx_path = "draft_from_colleague.docx",
  output_dir = "project/"
)

# What you get:
result$qmd_path         # QMD with content
result$references_json  # CSL-JSON for rendering
result$field_codes_json # Zotero metadata for round-trip
result$n_citations      # Number of citations found
```

Or extract citations only (without converting content):

```{r}
#| label: extract-only
extraction <- extract_citations(
  docx_path = "draft_from_colleague.docx",
  output_dir = "project/"
)
```

### Step 2: Edit in Quarto

Your QMD uses standard Pandoc citation syntax:

```markdown
Evidence suggests this approach is effective [@smith2023].
```

Edit freely—add text, reorganise sections, even add new citations.

### Step 3: Render with the docstyle extension

With the docstyle extension configured in `_quarto.yml`, render your document:

```bash
quarto render project/paper.qmd
```

The extension automatically:

1. Generates a `reference.docx` from your CSS + YAML configuration
2. Renders the QMD using Quarto/Pandoc
3. Injects Zotero field codes so citations remain editable in Word

## The "offline" advantage

You might ask: *Why not use the Better BibTeX Pandoc filter?*

The BBT filter (`zotero.lua`) is excellent, but requires that **every citation exists in your local Zotero library**.

docstyle's approach uses citation data **from the original Word document**:

- **No Zotero Desktop required**: Render without Zotero running
- **No missing references**: If your colleague cited a paper you don't have, docstyle preserves it exactly
- **No library sync issues**: Works offline with the extracted metadata

## Validation

Before round-tripping, validate that citations match your canonical library:

```{r}
#| label: validate
validate_citations(
  docx_path = "draft.docx",
  bib_path = "master_library.bib"
)
```

This reports:

- **Matches**: Citations correctly linked
- **Orphans**: Citations in the document missing from your library

## Adding new citations without a Word round-trip

If you are writing a document from scratch in Quarto (no existing Word file to harvest from), you can add citations directly from your local Zotero library:

```{r}
#| label: add-citations
library(docstyle)

# Add one or more citekeys from Zotero to _docstyle/field-codes.json
add_citations_from_zotero(c("tyas1998", "vangeli2011"))
```

This requires Zotero to be running with the [Better BibTeX](https://retorque.re/zotero-better-bibtex/) extension installed. Each citekey is looked up against your local Zotero library, and the CSL-JSON metadata plus Zotero URI are written into `_docstyle/field-codes.json`. By default, `references.bib` is also updated for LaTeX/Typst workflows.

```{r}
#| label: add-citations-options
# Suppress BibTeX export (docx-only workflow)
add_citations_from_zotero("tyas1998", write_bib = FALSE)

# Specify project directory explicitly
add_citations_from_zotero("tyas1998", project_dir = "path/to/project")
```

Behaviour:

- **Existing entries are never overwritten.** Re-running with the same citekey is safe.
- **Not-found citekeys produce a warning** and are skipped.
- **`write_bib = TRUE` (default)** calls `export_bibliography()` to sync `_docstyle/references.bib`.

After running, reference the citekey in your QMD with standard `[@citekey]` syntax and `quarto render` will inject a live `ADDIN ZOTERO_ITEM` field code into the output DOCX.

### Comparison with the round-trip workflow

| Scenario | Recommended approach |
|----------|---------------------|
| Starting from an existing Word file with citations | `docx_to_qmd()` (harvests citations automatically) |
| Writing from scratch, adding new references | `add_citations_from_zotero()` |
| Adding a citation to an existing project | `add_citations_from_zotero()` |
| Rendering without Zotero running | Either: metadata is cached in `field-codes.json` |

## Lower-level functions

For more control, use the individual functions:

### `extract_citations()`

Extract Zotero field codes from Word:

```{r}
#| label: extract-detailed
extraction <- extract_citations("document.docx", output_dir = "project/")
extraction$references_path   # Path to references.json
extraction$field_codes_path  # Path to field-codes.json
extraction$citations         # Parsed citation objects
```

### `inject_zotero_fields()`

Inject field codes into an existing Word document:

```{r}
#| label: inject
inject_zotero_fields(
  docx_path = "rendered.docx",
  field_codes_path = "field-codes.json"
)
```

## Troubleshooting

### Missing citations in output

Check that:

1. Citation keys in QMD match keys in `references.json`
2. `field-codes.json` contains entries for all citations
3. Run `validate_citations()` to identify mismatches

### Field codes not working in Word

Verify the output document:

1. Open in Word
2. Press Alt+F9 to show field codes
3. Zotero citations should show as `ADDIN ZOTERO_ITEM ...`

If they appear as static text, the injection may have failed. Check the console for warnings.

## Next steps

- **[Round-trip workflow](how-to-roundtrip.html)**: Complete workflow tutorial
- **[Getting started](tutorial-getting-started.html)**: docstyle basics
- **[CSS to Word](css-to-word.html)**: Customise document styling
